---
title: "Introduction 2 Data Analytics with Pokemon"
author: "Marta Fajlhauer"
date: "3 February 2018"
output:
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Before we begin

### Advantages of R:
* It is free to use

* It is open source: it encourages open collaboration. Eveyrone with sufficient skills can create a package to develop furter R. 

* It has a huge library of packages available to run every form of analysis. 

* R is flexible

* R has a nice visualisation graphics that we will see later in the course.

### Challenges of using R

You need to learn programming! ;p It is also fun

### Packages

R has many additional packages 

To use a package it needs to be installed. 

You only need to install a package once.

To use a package, you need to load the package each time you use R.


### Installation

Option 1. Use the install.packages function.

Note that some packages rely on other packages.

depencies = TRUE ensures that dependencies are also installed.

```{r install packages, eval=FALSE}
install.packages("dplyr", dependencies = T)
install.packages("VIM", dependencies = T)
install.packages("googleVis", dependencies = T)
install.packages("ggplot2", dependencies = T)
install.packages("scales", dependencies = T)
install.packages("ggridges", dependencies = T)
```

### Understanding directories
```{r directories}
# R has a working directory.
# This is important when loading and saving files to disk
getwd() # show the current working directory
setwd("C:/Users/fajlh/Documents/USB stick/statistics/Microsoft")
```

you can use setwd in R studio easily. Go to session -> set working directory
In the Files you can see all folders available in your working directory

Tips: 

* Try to avoid spaces in file names (use hyphen or underscore instead)

* If you do use spaces, then you'll need to escape the space with a slash  (e.g., ("my\ documents")

* Use backslash as the directory separator 

* Store all relevant files for a project within the project working directory

```{r loading libraries, message=FALSE, comment=FALSE, warning=FALSE}
library(dplyr) # main package for data shape and cleaning
library(VIM)
library(googleVis) # nice visual package
library(ggplot2) # main visual package
library(scales) # improve ggplot
library(ggridges) 
library(plotrix)
library(plotly) # Nice visual graphics
library(psych) # statsitics package
library(corrgram) # corrolelogram package
library(treemap)
```

# Getting the dataset

### Short information about data set

```{r Download data set, message=FALSE, comment=FALSE, warning=FALSE}
library(readr)
Pokemon <- read_csv("pokemon_alopez247.csv")
```

The dataset comes from https://www.kaggle.com/alopez247/pokemon/data. 
Metadata:
* Number. Pokémon ID in the Pokédex.

* Name. Name of the Pokémon.

* Total. Sum of all the base stats (Health Points, Attack, Defense, Special Attack, Special Defense, and Speed).

* HP. Base Health Points.

* Attack. Base Attack.

* Defense. Base Defense.

* Sp_Atk. Base Special Attack.

* Sp_Def. Base Special Defense.

* Speed. Base Speed.

* Generation. Number of the generation when the Pokémon was introduced.

* isLegendary. Boolean that indicates whether the Pokémon is Legendary or not.

* Color. Color of the Pokémon according to the Pokédex.

* hasGender. Boolean that indicates if the Pokémon can be classified as female or male.

* Pr_male. In case the Pokémon has Gender, the probability of its being male. The probability of being female is, of course, 1 minus this value.

* Egg_Group_1. Egg Group of the Pokémon. 

* Egg_Group_2. Second Egg Group of the Pokémon, in case it has two.

* hasMegaEvolution. Boolean that indicates whether the Pokémon is able to 

* Mega-evolve or not.

* Height_m. Height of the Pokémon, in meters.

* Weight_kg. Weight of the Pokémon, in kilograms.

* Catch_Rate. Catch Rate.

* Body_Style. Body Style of the Pokémon according to the Pokédex.

# Basic Arithmetic and Logical Operations
```{r basic_multiplication}
# You can use R like a calculator
1 + 5 # addition
10 - 7 # subtraction
10 * 10 # multiplcation
90 / 9 # division
11 ^ 2 # exponentiation

abs(-10) # absolute value
ceiling(2.75) # round up to next integer
floor(3.25) # round down to next integer
sqrt(100) # square roots
exp(2) # exponents
pi # mathematical constant pi
exp(1) # mathematical constant e
log(100) # natural logs (i.e., base e)
log(100, base= 10) # base 10 logs


# Use parentheses to clarify order of operations
(1 + 1 ) * 2
1 + (1 * 2)

# You can test for equality
# TRUE and FALSE are keywords
# T and F are synonyms
TRUE
FALSE
1 == 2 # Equality (Return TRUE if equal)
1 != 2 # Inequality (Return FALSE if unequal)
10 > 9 # Greater than
9 < 10 # Less than
10 <= 10 # Less than or equal
2 %in% c(1, 2 ,3) # is the number in the vector
!(5>1)
result <- 4<5
result2 <- !(5>1) #is not (TRUE)
result | result2 # result 1 or result 2 (because one of them is TRUE)
result & result2 # result 1 and result 2 (because one is not true)
isTRUE(result2)

# TRUE and FALSE coerces to 1 and 0 respectively
as.numeric(TRUE)
as.numeric(FALSE)

# Logical converting to 0, 1 is useful
x <- c(2, 5 ,7 ,10, 15)
x > 5
sum(x > 5)  # sum of a 0-1 variable is a count
mean(x > 5) # mean of a 0-1 variable is a proportion

```

# Basic language features
Assignment:
To assign values to a variable either use "<-" or "="
"<-" is the more common convention R
```{r basic language features, error = TRUE}
x <- 1 + 1
x
```

### Type of variables.

There are different ypes of variables and they are important in selection of visualisation technique and decision of correct statistical modelling performed on the dataset. 

```{r type of variables}
typeof(Pokemon$Name) # character variable
typeof(Pokemon$Total) # numerical variable
```

### Variable name rules
1. Variable names generally
+ Start with a letter (lower or uppercase)
+ Followed by letters, numbers, underscore (_), or period (.)
+ No spaces

These do not work
```{r, error=TRUE}
my variable <- 1234
1234variable <- 1234
1234 <- 1234
```

This works
```{r, error=TRUE}
myvariable <- 1234
my_variable <- 1234
my_variable <- 1234
myvariable123 <- 1234
myVariable <- 1234
my.variable <- 1234
```
R has many naming conventions
As a matter of preference, style, and convenience, I prefer:
1. Short but descriptive names

* Less than 8 characters for names of lists and data.frames

* Less than 15 characters for variables names in data.frames

2. Use underscore to separate words within a variable name

3. Avoid upper case letters


Names starting with a period are hidden
```{r, error=TRUE}
.myvariable <- 1234
ls()
ls(all.names = TRUE)
```

R is case sensitive
```{r, error=TRUE}
test <- "lower case"
TEST <- "upper case"
TEST 
test # The original value was not lost
     # because test is different to TEST
Test # This variable does not exist

Test <- "title case"
Test
```

### Data Frames:
* Data frames are the standard data strucure used for storing data. 

* Columns can be of different data types (e.g., character, numeric, logical, etc.)

### Tip: 
* Some functions work with matrices, some work with data.frames, and some work with both.
* If you are wanting to store data like you might store in a database, then you'll generaly want a data.frame.
* If you are dealing with a mathematical object that you you want to perform a mathematical operation on, then you generally want a matrix (e.g., correlation matrix, covariance matrix, distance matrix in MDS, matrices used for matrix algebra).

### Extracting observations (i.e., rows) and variables (i.e., columns).
There are similarities to matrices and lists
Select observations
```{r extracting observations}
Pokemon[1:5, ] # by row number
Pokemon[c(5,4,3,2,1), ] # re-order
Pokemon[Pokemon$isLegendary == "True", ] # by logical vector
Pokemon[c("10", "56"), ] # by rownames

# Select variables 
Pokemon[, c(1,2)] # by position like a matrix
Pokemon[c(1,2)] # by position like a list
Pokemon[ ,c("Name", "Type_1")] # by name like a matrix
Pokemon[c("Name", "Type_1")] # by name like a list
```

```{r, eval=FALSE}
Pokemon$Total # by name to get a single variable
```

### Names
```{r names}
names(Pokemon) # get variable names
colnames(Pokemon) # but this also works
```

```{r, eval=FALSE}
rownames(Pokemon) # rows can also have names
```
#### Tip: 
Avoid row names. Add another variable to the data.frame to store this information.


### Examine first few rows
```{r looking at the data}
head(Pokemon) # first 6 rows
head(Pokemon, n = 10) # first 7 rows
tail(Pokemon) # last few rows
View(Pokemon) # Rstudio function to open data in viewer or click on the icon in the Environment pane

# How many rows and columns?
dim(Pokemon) # rows and column counts
nrow(Pokemon) # row count
ncol(Pokemon) # column count
```

### Basics statistics on dataset


```{r Basics statistics}
str(Pokemon)
summary(Pokemon)
describe(CrimeData$PctBornSameState)
```

Function str in Basics R shows us structure of the dataset. We see that there are 23 variables and 721 observations. There is 10 character features and rest numerical. Summary statistics shows us 5 point statistics for numerical features. We can nicely visualise missing numbers by:

```{r missing numbers visualisation}
aggr_plot <- aggr(Pokemon, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(data), cex.axis=.5, gap=3, ylab=c("Histogram of missing data","Pattern"))
```

As our main concern is not to run advanced statistical analyses but we want to explore R we will simply remove the variables with missing numbers. 

```{r removing variables with NA}
myvars <- names(Pokemon) %in% c("Egg_Group_2", "Type_2", "Pr_Male")
Pokemon <- Pokemon[!myvars]
```

# Exploratory Data Analysis so let's do a lot of Nice, Colourful Graphs. 

### Histogram: 
A basic histogram bins a variable into fixed width buckets and returns the number of data points that falls into each bucket. 
The primary disadvantge of histograms is that you must decide ahead of time how wide the buckets are. 

```{r histogram basic}
hist(Pokemon$Total, binw = 20) # histogram
```

It is so easy to do histogram using ggplot2 and the package allows to do much nicer and more colourful graphs as we will see. 
```{r basic gghist}
p3 <- ggplot(data=Pokemon, aes(x=Total))
range(Pokemon$Total) # tip: always check range of the variable to estimate binwidth
p3 + geom_histogram(binwidth =20) # set up the binwidth

```

We can colour the histogram using another colour as second variable.  
```{r colour histogram in depends of another variable}
p3 + geom_histogram(binwidth =20, aes(fill = Type_1)) # map second variable as a colour
```

We can add the black line to differentiate easier colours of the variables. 
```{r adding black line}
p3 <- p3 + geom_histogram(binwidth =20, aes(fill = Type_1), colour = 'Black') # set up border for different types in the histogram
p3
```

Having graph ready we would like to add labels so we can include this into our project. 
```{r adding text}
p3 + 
  xlab('Type') + # add the x axis title
  ylab('Total score for Pokemons') + # add the y-axis title
  ggtitle('Total score for different types') + # add the title
  theme(axis.title.x = element_text(colour = 'Blue', size = 20), # change colour and size of x-axis title
        axis.title.y = element_text(colour = 'Red', size = 20), # change colour and size of y-axis title
        legend.title = element_text(size = 20), # change the size of title in the legend
        legend.text = element_text(size=10), # change the size of content in the legend
        legend.position = c(1,1), # according to: 0,0 = origin, 0,1 = on the y-axis where x=0, 1,0 = on the x-axis where y=0
        legend.justification = c(1,1),
        plot.title = element_text(colour = 'DarkGreen', size = 30)) # which corner of the text box you want to put in which corner of the grid box.
```

### distribution
You can think of a 'density plot' as a "continuous histogram" of a variable, except the area under the density plot is equal to 1. 
You should use a log scale to better visualise data that is heavily skewed.

```{r basic density}
plot(density(Pokemon$HP)) # density plot
```

Again ggplot2 allows us to produce much nicer graph very quickly. 
```{r basic ggdist}
p3 <- ggplot(data=Pokemon, aes(x=HP))
p3 + geom_density() # create density plot
```

We can see how densities will differ according to what groups it belongs to. 
```{r imroving density}
p3 + geom_density(aes(fill = Type_1)) # create density plot for different types
```

Adding labels and black line allows us to read the graph much easier. 
```{r}
p3 <- p3 + geom_density(aes(fill = Type_1), position = 'stack') # so you can see better density distribution for different types
p3
p3 + 
  xlab('Type') + # add the x axis title
  ylab('Density of scores') + # add the y-axis title
  ggtitle('Density plots for different types') + # add the title
  theme(axis.title.x = element_text(colour = 'Blue', size = 20), # change colour and size of x-axis title
        axis.title.y = element_text(colour = 'Red', size = 20), # change colour and size of y-axis title
        legend.title = element_text(size = 20), # change the size of title in the legend
        legend.text = element_text(size=10), # change the size of content in the legend
        legend.position = c(1,1), # according to: 0,0 = origin, 0,1 = on the y-axis where x=0, 1,0 = on the x-axis where y=0
        legend.justification = c(1,1),
        plot.title = element_text(colour = 'DarkGreen', size = 30)) # which corner of the text box you want to put in which corner of the grid box.
```

We can also find that legendary pokemons have higher total skills than non-legendary and that the legendary ones were: water, steel, rock, psychic, fire, electric, dragon. 
```{r}
ggplot(Pokemon, aes(x = Total, y = Type_1, fill = isLegendary)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = c(-0.2,-0.1))
```

### Box plot
A bar chart is a histogram for discrete data. 
The simplest one we can create using basic R
```{r basic boxplot}
boxplot(Pokemon$Speed) # box plot
```

We can check how distribution vary according to group.
```{r basic boxplot by group}
boxplot(Pokemon$Speed~ Pokemon$Type_1) # box plot by group
```
 Again we would prefer ggplot2 instead of base R
```{r basic ggbox}
p5 <- ggplot(data = Pokemon, aes(x = Type_1, y=Speed, colour=Type_1))
p5 + geom_boxplot()
```

We can improve the boxplots adding points for the dataset
```{r improving ggbox}
p5 + geom_boxplot(size = 1.1) + geom_point() # too pretty to be good
```

But this is too nice to be good so we need to jiter this a little. 
```{r last touch}
p5 + geom_boxplot(size = 1.3) + geom_jitter() # it's just randomly thrown points helping you too see the data better
p5 + geom_jitter() + geom_boxplot(size = 1.3, alpha = 0.5) # boxplot on top and set up transparency
```

However my favourite one would be using plotly. 
```{r plotly}
p <- plot_ly(Pokemon, y = ~Speed, color= ~Type_1, type = "box")
p
```

### Bar plot
These total base stats include average health points

```{r circular barplot}
ggplot(Pokemon, aes(x = Type_1, y = HP ,fill = Type_1)) + 
geom_bar(width = 0.85, stat="identity") +
# To use a polar plot and not a basic barplot
coord_polar(theta = "y") +
#Remove useless labels of axis
xlab("") + ylab("") +
#Increase ylim to avoid having a complete circle
#Add group labels close to the bars :
geom_text(data = Pokemon, hjust = 1, size = 3, aes(x = Type_1, y = 0, label = Type_1)) +
#Remove useless legend, y axis ticks and y axis text
theme(legend.position = "none" , axis.text.y = element_blank() , axis.ticks = element_blank())
```


### Line plots
Works best when the relationship between two variables is relatively clean: each x value has a unique y value. When the data is not so cleanle related you'll want to use scatterplot instead

### scatterplot

Scatterplots allows us to see relationship between features in the dataset. If one increases is the other one increasing/decreasing?

```{r scatterplot}
par(mfrow=c(1,1)) # create a layout of graphic wondow, useful if you want to put more than one graph in one graphic window 
plot(Pokemon$Total, Pokemon$Attack) # plot scatterplot for the variables
plot(Pokemon$Total) # plot one variable only
library(psych)
pairs.panels(Pokemon[,c("Total", "HP", "Attack", "Defense", 'Speed')], 
             density = TRUE,
             ellipses = T, 
             lm = T, 
             method = 'spearman',
             rug = F,
             smoother = F)
```

We can produce bubble plot where ssize of the bubble depends on one variable, in this case total score, and we plot relationship between HP and Attack. 
```{r bubble plot, results='asis', tidy=FALSE}
Bubble <- gvisBubbleChart(Pokemon, idvar="Type_1", 
                          xvar="HP", yvar="Attack",
                          colorvar="Generation", sizevar="Total", 
                          options = list(
                            width=2500,
                            height=1500
                          ))
plot(Bubble)
```

Plotly produces nice plot as well. 
```{r plotly}
p <- plot_ly(
  Pokemon, x = ~HP, y = ~Attack,
  # Hover text:
  text = ~paste("Attack: ", Attack, '<br>Generation:', Generation),
  color = ~Generation, size = ~HP
)
p
```

### Bar chart 


```{r barplot horizontal }
ggplot(data = Pokemon, aes(x=Type_1)) + # Plot bar chart as before
  geom_bar() +
  coord_flip() + # Flip the x and y axis
  theme(axis.text.y = element_text(size=rel(0.8))) #reduce the size of the y-axis tick labels 
```

```{r improved}
TypeSum <- table(Pokemon$Type_1) # aggregate the data by state of residence
TypeOf <- as.data.frame(TypeSum) # convert to data frame
colnames(TypeOf) <- c('Type', 'count') # change the column names
summary(TypeOf) # default ordering is alphabetical
TypeOf <- transform(TypeOf, Type = reorder(Type, count)) # reorder
summary(TypeOf) # The region is now count ordered
ggplot(TypeOf) +
  geom_bar(aes(x=Type, y=count), 
           stat = 'identity',
           fill = 'blue') +
  coord_flip() +
  theme(axis.text.y = element_text(size = rel(0.8)))
```

### Pie chart
```{r Pie Chart, warning=FALSE}
AttackSkills <- Pokemon[,c('Type_1' ,'Attack')]
AttackSkills <- aggregate(AttackSkills, by = list(Type = AttackSkills$Type_1), FUN = 'mean')
AttackSkills <- AttackSkills[,c(1,3)]
AttackSkills <- AttackSkills[order(AttackSkills$Attack),]
slices <- AttackSkills$Attack 
lbls <- AttackSkills$Type
pie3D(slices,labels=lbls,explode=0.1,
  	main="Pie Chart of Types")
```
```{r PieVis}
Pie <- gvisPieChart(AttackSkills, options = list(
  width=2500,
  height=1500
  ))
plot(Pie)
```

All pokemons have a total sum of all the base stats (Health Points, Attack, Defense, Special Attack, Special Defense, and Speed) as a ranking. Those with the higest ranking are the strongest ones.
```{r ranking of total skills, warning=FALSE, eval=FALSE}
TotalSkills <- Pokemon[,c('Type_1' ,'Total')]
TotalSkills <- aggregate(TotalSkills, by = list(Type = TotalSkills$Type_1), FUN = 'mean')
TotalSkills <- TotalSkills[,c(1,3)]
TotalSkills <- TotalSkills[order(TotalSkills$Total),]
Gauge <-  gvisGauge(TotalSkills, 
                    options=list(min=365, max=501, 
                                 greenFrom=457,
                                 greenTo=501, 
                                 yellowFrom=411, 
                                 yellowTo=457,
                                 redFrom=365, 
                                 redTo=411,
                                 width = 1000,
                                 height = 900))
plot(Gauge)
```



```{r defense skills map, warning=FALSE}
library(treemap)
treemap(Pokemon, #Your data frame object
        index=c("Type_1","Name"),  #A list of your categorical variables
        vSize = "Defense",  #This is your quantitative variable
        type="index", #Type sets the organization and color scheme of your treemap
        palette = "Reds",  #Select your color palette from the RColorBrewer presets or make your own.
        title="Defense skills of Pokemon", #Customize your title
        fontsize.title = 14) #Change the font size of the title)
```

```{r corrolelogram}
library(corrgram)
Num <- Pokemon[,c("Total", "HP", "Attack", "Defense", "Sp_Atk", "Sp_Def","Speed")]
corrgram(Num, upper.panel=panel.pie, diag.panel = panel.density, lower.panel = panel.cor,main="Pokemon attributes",
         col.regions = colorRampPalette(c("yellow", "blue", "red", "violet")))
```
# Exporting data 
```{r exporting data}

# Interal R format
# Good option if you need to re-open data in R
save(Pokemon, file="Pokemon.rdata")


# csv
# Good option if you need to get data into other software
# This should open in almost all other software (e.g. Excel, SPSS, etc.)
write.csv(Pokemon, file = "Pokemon.csv", row.names = FALSE) # exclude row.names

# If you need more flexibility in terms of delimiters, etc.
write.table(Pokemon, file = "Pokemon.tsv", sep = "\t") # e.g., tab delimiter
```

# Questions

```{r Exercises 1}
# To answer on these questions use dataset diamonds in the ggplot2 package
library(ggplot2)
# 1. How many observations are there?

# 2. Show the first 7 rows of the diamonds dataset

# 3. show the structure of diamonds using str function

# 4. How many rows is in the dataset

# 5. what kind of variable is color

# 6. Using summary function tell how many diamonds has ideal cut

# 7. Extract data only for diamonds that cost more than $2401 and store as Expensive

# 8. Using describe function inside psych tell what is mean, standard deviation and range of price for the diamonds.

# 9. Find a range of carat for expensive diamonds using basic r

# 10. Create a histogram using basic r for the carats using binwith 0.5

# 11. Create a ggplot using ggplot2 library for the carats store them as carat

# 12. Create histogram and add the binwidth the same as we used in basic r

# 13. colour the histogram mapping cut

# 14. set up border for different types in the histogram

# 15. Add the main title to the graph

# 16. Create a density plot using basic r for the price

# 17. Create a ggplot using ggplot2 library for the price and store it as price

# 18. Create density plot

# 19. Colour the density plot for different cuts

# 20. Use position stack to see better density distribution for different cuts

```

```{r Exercises 1}
# 1. How many observations are there?
dim(diamonds)
# 2. Show the first 7 rows of the diamonds dataset
head(diamonds, n = 7)
# 3. show the structure of diamonds using str function
str(diamonds)
# 4. How many rows is in the dataset
nrow(diamonds)
# 5. what kind of variable is color
typeof(diamonds$color) 
# 6. Using summary function tell how many diamonds has ideal cut
summary(diamonds$cut)
# 7. Extract data only for diamonds that cost more than $2401 and store as Expensive
Expensive <- diamonds[diamonds$price > 2401,]
# 8. Using describe function inside psych tell what is mean, standard deviation and range of price for the diamonds.
describe(Expensive$price)
# 9. Find a range of carat for expensive diamonds using basic r
range(Expensive$carat)
# 10. Create a histogram using basic r for the carats using binwith 0.5
hist(Expensive$carat, binw = 0.5)
# 11. Create a ggplot using ggplot2 library for the carats store them as carat
carat <- ggplot(data = Expensive, aes(x=carat))
carat
# 12. Create histogram and add the binwidth the same as we used in basic r
carat + geom_histogram(binwidth = 0.5)
# 13. colour the histogram mapping cut
carat + geom_histogram(binwidth = 0.5, aes(fill = cut))
# 14. set up border for different types in the histogram
carat + geom_histogram(binwidth = 0.5, aes(fill = cut), colour = 'Black')
# 15. Add the main title to the graph
carat + geom_histogram(binwidth = 0.5, aes(fill = cut), colour = 'Black') +
  ggtitle('Carats of expensive diamonds')
# 16. Create a density plot using basic r for the price
plot(density(Expensive$price))
# 17. Create a ggplot using ggplot2 library for the price and store it as price
Price <- ggplot(data = Expensive, aes(x=price))
# 18. Create density plot
Price + geom_density() # create density plot
# 19. Colour the density plot for different cuts
Price + geom_density(aes(fill = cut)) # create density plot for different cuts
# 20. Use position stack to see better density distribution for different cuts
Price + geom_density(aes(fill = cut), position = 'stack')
```


