---
title: "Introduction 2 Data Analytics with Pokemon"
author: "Marta Fajlhauer"
date: "3 February 2018"
output:
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Before we begin

### Advantages of R:
* It is free to use

* It is open source: it encourages open collaboration. Eveyrone with sufficient skills can create a package to develop furter R. 

* It has a huge library of packages available to run every form of analysis. 

* R is flexible

* R has a nice visualisation graphics that we will see later in the course.

### Challenges of using R

You need to learn programming! ;p It is also fun

### Packages

R has many additional packages 

To use a package it needs to be installed. 

You only need to install a package once.

To use a package, you need to load the package each time you use R.


### Installation

Option 1. Use the install.packages function.

Note that some packages rely on other packages.

depencies = TRUE ensures that dependencies are also installed.

```{r install packages, eval=FALSE}
install.packages("dplyr", dependencies = T)
install.packages("VIM", dependencies = T)
install.packages("googleVis", dependencies = T)
install.packages("ggplot2", dependencies = T)
install.packages("scales", dependencies = T)
install.packages("ggridges", dependencies = T)
```

### Understanding directories
```{r directories}
# R has a working directory.
# This is important when loading and saving files to disk
getwd() # show the current working directory
setwd("C:/Users/fajlh/Documents/USB stick/statistics/Microsoft")
```

you can use setwd in R studio easily. Go to session -> set working directory
In the Files you can see all folders available in your working directory

Tips: 

* Try to avoid spaces in file names (use hyphen or underscore instead)

* If you do use spaces, then you'll need to escape the space with a slash  (e.g., ("my\ documents")

* Use backslash as the directory separator 

* Store all relevant files for a project within the project working directory

```{r loading libraries, message=FALSE, comment=FALSE, warning=FALSE}
library(dplyr)
library(VIM)
library(googleVis)
library(ggplot2)
library(scales)
library(ggridges)
library(plotrix)
library(plotly)
library(psych)
```

# Getting the dataset

### Short information about data set

```{r Download data set, message=FALSE, comment=FALSE, warning=FALSE}
library(readr)
Pokemon <- read_csv("pokemon_alopez247.csv")
```

The dataset comes from https://www.kaggle.com/alopez247/pokemon/data. 
Metadata:
* Number. Pokémon ID in the Pokédex.

* Name. Name of the Pokémon.

* Total. Sum of all the base stats (Health Points, Attack, Defense, Special Attack, Special Defense, and Speed).

* HP. Base Health Points.

* Attack. Base Attack.

* Defense. Base Defense.

* Sp_Atk. Base Special Attack.

* Sp_Def. Base Special Defense.

* Speed. Base Speed.

* Generation. Number of the generation when the Pokémon was introduced.

* isLegendary. Boolean that indicates whether the Pokémon is Legendary or not.

* Color. Color of the Pokémon according to the Pokédex.

* hasGender. Boolean that indicates if the Pokémon can be classified as female or male.

* Pr_male. In case the Pokémon has Gender, the probability of its being male. The probability of being female is, of course, 1 minus this value.

* Egg_Group_1. Egg Group of the Pokémon. 

* Egg_Group_2. Second Egg Group of the Pokémon, in case it has two.

* hasMegaEvolution. Boolean that indicates whether the Pokémon is able to 

* Mega-evolve or not.

* Height_m. Height of the Pokémon, in meters.

* Weight_kg. Weight of the Pokémon, in kilograms.

* Catch_Rate. Catch Rate.

* Body_Style. Body Style of the Pokémon according to the Pokédex.

### Type of variables.

There are different ypes of variables and they are important in selection of visualisation technique and decision of correct statistical modelling performed on the dataset. 

```{r type of variables}
typeof(Pokemon$Name) # character variable
typeof(Pokemon$Total) # numerical variable
```

### Data Frames:
* Data frames are the standard data strucure used for storing data. If you have used other software (e.g., SPSS, Excel, etc.), this is what you may think of as a "dataset".
* Columns can be of different data types (e.g., character, numeric, logical, etc.)

### Tip: 
* Some functions work with matrices, some work with data.frames, and some work with both.
* If you are wanting to store data like you might store in a database, then you'll generaly want a data.frame.
* If you are dealing with a mathematical object that you you want to perform a mathematical operation on, then you generally want a matrix (e.g., correlation matrix, covariance matrix, distance matrix in MDS, matrices used for matrix algebra).

### Extracting observations (i.e., rows) and variables (i.e., columns).
There are similarities to matrices and lists
Select observations
```{r extracting observations}
Pokemon[1:5, ] # by row number
Pokemon[c(5,4,3,2,1), ] # re-order
Pokemon[Pokemon$isLegendary == "True", ] # by logical vector
Pokemon[c("10", "56"), ] # by rownames

# Select variables 
Pokemon[, c(1,2)] # by position like a matrix
Pokemon[c(1,2)] # by position like a list
Pokemon[ ,c("Name", "Type_1")] # by name like a matrix
Pokemon[c("Name", "Type_1")] # by name like a list
```

```{r, eval=FALSE}
Pokemon$Total # by name to get a single variable
```

### Names
```{r names}
names(Pokemon) # get variable names
colnames(Pokemon) # but this also works
```

```{r, eval=FALSE}
rownames(Pokemon) # rows can also have names
```
#### Tip: 
Avoid row names. Add another variable to the data.frame to store this information.


### Examine first few rows
```{r looking at the data}
head(Pokemon) # first 6 rows
head(Pokemon, n = 10) # first 7 rows
tail(Pokemon) # last few rows
View(Pokemon) # Rstudio function to open data in viewer or click on the icon in the Environment pane

# How many rows and columns?
dim(Pokemon) # rows and column counts
nrow(Pokemon) # row count
ncol(Pokemon) # column count
```

### Basics statistics on dataset


```{r Basics statistics}
str(Pokemon)
summary(Pokemon)
describe(CrimeData$PctBornSameState)
```

Function str in Basics R shows us structure of the dataset. We see that there are 23 variables and 721 observations. There is 10 character features and rest numerical. Summary statistics shows us 5 point statistics for numerical features. We can nicely visualise missing numbers by:

```{r missing numbers visualisation}
aggr_plot <- aggr(Pokemon, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(data), cex.axis=.5, gap=3, ylab=c("Histogram of missing data","Pattern"))
```

As our main concern is not to run advanced statistical analyses but we want to explore R we will simply remove the variables with missing numbers. 

```{r removing variables with NA}
myvars <- names(Pokemon) %in% c("Egg_Group_2", "Type_2", "Pr_Male")
Pokemon <- Pokemon[!myvars]
```

# Exploratory Data Analysis so let's do a lot of Nice, Colourful Graphs. 

### Histogram: 
A basic histogram bins a variable into fixed width buckets and returns the number of data points that falls into each bucket. 
The primary disadvantge of histograms is that you must decide ahead of time how wide the buckets are. 

```{r histogram}
hist(Pokemon$Total, binw = 20) # histogram
p3 <- ggplot(data=Pokemon, aes(x=Total))
range(Pokemon$Total) # tip: always check range of the variable to estimate binwidth
p3 + geom_histogram(binwidth =20) # set up the binwidth
p3 + geom_histogram(binwidth =20, aes(fill = Type_1)) # map second variable as a colour
p3 + geom_histogram(binwidth =20, aes(fill = Type_1), colour = 'Black') # set up border for different types in the histogram
r <- p3 + geom_histogram(binwidth =20, aes(fill = Type_1), colour = 'Black')
r + 
  xlab('Type') + # add the x axis title
  ylab('Total score for Pokemons') + # add the y-axis title
  ggtitle('Total score for different types') + # add the title
  theme(axis.title.x = element_text(colour = 'Blue', size = 20), # change colour and size of x-axis title
        axis.title.y = element_text(colour = 'Red', size = 20), # change colour and size of y-axis title
        legend.title = element_text(size = 20), # change the size of title in the legend
        legend.text = element_text(size=10), # change the size of content in the legend
        legend.position = c(1,1), # according to: 0,0 = origin, 0,1 = on the y-axis where x=0, 1,0 = on the x-axis where y=0
        legend.justification = c(1,1),
        plot.title = element_text(colour = 'DarkGreen', size = 30)) # which corner of the text box you want to put in which corner of the grid box.
```

### distribution
You can think of a 'density plot' as a "continuous histogram" of a variable, except the area under the density plot is equal to 1. 
You should use a log scale to better visualise data that is heavily skewed.

```{r density}
plot(density(Pokemon$HP)) # density plot
p3 <- ggplot(data=Pokemon, aes(x=HP))
p3 + geom_density() # create density plot
p3 + geom_density(aes(fill = Type_1)) # create density plot for different types
p3 + geom_density(aes(fill = Type_1), position = 'stack') # so you can see better density distribution for different types
```

We can also find that legendary pokemons have higher total skills than non-legendary and taht the legendary ones were: water, steel, rock, psychic, fire, electric, dragon. 
```{r}
ggplot(Pokemon, aes(x = Total, y = Type_1, fill = isLegendary)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```

### Box plot
A bar chart is a histogram for discrete data. 

```{r boxplot}
boxplot(Pokemon$Speed) # box plot
boxplot(Pokemon$Speed~ Pokemon$Type_1) # box plot by group
p5 <- ggplot(data = Pokemon, aes(x = Type_1, y=Speed, colour=Type_1))
p5 + geom_boxplot()
p5 + geom_boxplot(size = 1.1) # increase size of the borders
p5 + geom_boxplot(size = 1.1) + geom_point() # too pretty to be good
p5 + geom_boxplot(size = 1.3) + geom_jitter() # it's just randomly thrown points helping you too see the data better
p5 + geom_jitter() + geom_boxplot(size = 1.3, alpha = 0.5) # boxplot on top and set up transparency
```

```{r plotly}
p <- plot_ly(Pokemon, y = ~Speed, color= ~Type_1, type = "box")
p
```
### Violin plot

A way of visualisation of distribution of numerical variable that allows a deeper understanding of the density as density plot is drawn on sides of the violin

```{r violin plot}
ggplot(Pokemon, aes(factor(Type_1), Attack)) + 
  geom_violin(aes(fill = Type_1))
```

### Bar plot
These total base stats include average health points

```{r circular barplot}
ggplot(Pokemon, aes(x = Type_1, y = HP ,fill = Type_1)) + 
geom_bar(width = 0.85, stat="identity") +
# To use a polar plot and not a basic barplot
coord_polar(theta = "y") +
#Remove useless labels of axis
xlab("") + ylab("") +
#Increase ylim to avoid having a complete circle
#Add group labels close to the bars :
geom_text(data = Pokemon, hjust = 1, size = 3, aes(x = Type_1, y = 0, label = Type_1)) +
#Remove useless legend, y axis ticks and y axis text
theme(legend.position = "none" , axis.text.y = element_blank() , axis.ticks = element_blank())
```


### Line plots
Works best when the relationship between two variables is relatively clean: each x value has a unique y value. When the data is not so cleanle related you'll want to use scatterplot instead

### scatterplot

```{r scatterplot}
par(mfrow=c(1,1)) # create a layout of graphic wondow, useful if you want to put more than one graph in one graphic window 
plot(Pokemon$Total, Pokemon$Attack) # plot scatterplot for the variables
plot(Pokemon$Total) # plot one variable only
library(psych)
pairs.panels(Pokemon[,c("Total", "HP", "Attack", "Defense", 'Speed')], 
             density = TRUE,
             ellipses = T, 
             lm = T, 
             method = 'spearman',
             rug = F,
             smoother = F)
```

```{r bubble plot, results='asis', tidy=FALSE}
Bubble <- gvisBubbleChart(Pokemon, idvar="Type_1", 
                          xvar="HP", yvar="Attack",
                          colorvar="Generation", sizevar="Total", 
                          options = list(
                            width=2500,
                            height=1500
                          ))
plot(Bubble)
```

```{r plotly}
p <- plot_ly(
  Pokemon, x = ~HP, y = ~Attack,
  # Hover text:
  text = ~paste("Attack: ", Attack, '<br>Generation:', Generation),
  color = ~Generation, size = ~HP
)
p
```
### categorical variable 

```{r categorical variable}
# categorical variables
plot(Pokemon$Generation) # categorical variable will be presented in bar chart
plot(prop.table(table(Pokemon$Generation))) #p.m.f. of categorical variable
```

### Pie chart
There are different kind of Pokemons
```{r circular barplot}
ggplot(data = Pokemon, aes(x=Type_1)) + # Plot bar chart as before
  geom_bar() +
  coord_flip() + # Flip the x and y axis
  theme(axis.text.y = element_text(size=rel(0.8))) #reduce the size of the y-axis tick labels 
TypeSum <- table(Pokemon$Type_1) # aggregate the data by state of residence
TypeOf <- as.data.frame(TypeSum) # convert to data frame
colnames(TypeOf) <- c('Type', 'count') # change the column names
summary(TypeOf) # default ordering is alphabetical
TypeOf <- transform(TypeOf, Type = reorder(Type, count)) # reorder
summary(TypeOf) # The region is now count ordered
ggplot(TypeOf) +
  geom_bar(aes(x=Type, y=count), 
           stat = 'identity',
           fill = 'blue') +
  coord_flip() +
  theme(axis.text.y = element_text(size = rel(0.8)))
```

```{r Pie Chart, warning=FALSE}
AttackSkills <- Pokemon[,c('Type_1' ,'Attack')]
AttackSkills <- aggregate(AttackSkills, by = list(Type = AttackSkills$Type_1), FUN = 'mean')
AttackSkills <- AttackSkills[,c(1,3)]
AttackSkills <- AttackSkills[order(AttackSkills$Attack),]
slices <- AttackSkills$Attack 
lbls <- AttackSkills$Type
pie3D(slices,labels=lbls,explode=0.1,
  	main="Pie Chart of Types")
Pie <- gvisPieChart(AttackSkills, options = list(
  width=2500,
  height=1500
  ))
plot(Pie)
```

All pokemons have a total sum of all the base stats (Health Points, Attack, Defense, Special Attack, Special Defense, and Speed) as a ranking. Those with the higest ranking are the strongest ones.
```{r ranking of total skills, warning=FALSE, eval=FALSE}
TotalSkills <- Pokemon[,c('Type_1' ,'Total')]
TotalSkills <- aggregate(TotalSkills, by = list(Type = TotalSkills$Type_1), FUN = 'mean')
TotalSkills <- TotalSkills[,c(1,3)]
TotalSkills <- TotalSkills[order(TotalSkills$Total),]
Gauge <-  gvisGauge(TotalSkills, 
                    options=list(min=365, max=501, 
                                 greenFrom=457,
                                 greenTo=501, 
                                 yellowFrom=411, 
                                 yellowTo=457,
                                 redFrom=365, 
                                 redTo=411,
                                 width = 1000,
                                 height = 900))
plot(Gauge)
```



```{r defense skills map, warning=FALSE}
library(treemap)
treemap(Pokemon, #Your data frame object
        index=c("Type_1","Name"),  #A list of your categorical variables
        vSize = "Defense",  #This is your quantitative variable
        type="index", #Type sets the organization and color scheme of your treemap
        palette = "Reds",  #Select your color palette from the RColorBrewer presets or make your own.
        title="Defense skills of Pokemon", #Customize your title
        fontsize.title = 14) #Change the font size of the title)
```

```{r corrolelogram}
library(corrgram)
Num <- Pokemon[,c("Total", "HP", "Attack", "Defense", "Sp_Atk", "Sp_Def","Speed")]
corrgram(Num, upper.panel=panel.pie, diag.panel = panel.density, lower.panel = panel.cor,main="Pokemon attributes",
         col.regions = colorRampPalette(c("yellow", "blue", "red", "violet")))
```
# Exporting data 
```{r exporting data}

# Interal R format
# Good option if you need to re-open data in R
save(Pokemon, file="Pokemon.rdata")


# csv
# Good option if you need to get data into other software
# This should open in almost all other software (e.g. Excel, SPSS, etc.)
write.csv(Pokemon, file = "Pokemon.csv", row.names = FALSE) # exclude row.names

# If you need more flexibility in terms of delimiters, etc.
write.table(Pokemon, file = "Pokemon.tsv", sep = "\t") # e.g., tab delimiter
```

# Questions

```{r Exercises 1}
# 1. How many observations are there?

# 2. Show the first 7 rows of the diamonds dataset

# 3. show the structure of diamonds using str function

# 4. How many rows is in the dataset

# 5. what kind of variable is color

# 6. Using summary function tell how many diamonds has ideal cut

# 7. Extract data only for diamonds that cost more than $2401 and store as Expensive

# 8. Using describe function inside psych tell what is mean, standard deviation and range of price for the diamonds.

# 9. Find a range of carat for expensive diamonds using basic r

# 10. Create a histogram using basic r for the carats using binwith 0.5

# 11. Create a ggplot using ggplot2 library for the carats store them as carat

# 12. Create histogram and add the binwidth the same as we used in basic r

# 13. colour the histogram mapping cut

# 14. set up border for different types in the histogram

# 15. Add the main title to the graph

# 16. Create a density plot using basic r for the price

# 17. Create a ggplot using ggplot2 library for the price and store it as price

# 18. Create density plot

# 19. Colour the density plot for different cuts

# 20. Use position stack to see better density distribution for different cuts

```

```{r Exercises 1}
# 1. How many observations are there?
dim(diamonds)
# 2. Show the first 7 rows of the diamonds dataset
head(diamonds, n = 7)
# 3. show the structure of diamonds using str function
str(diamonds)
# 4. How many rows is in the dataset
nrow(diamonds)
# 5. what kind of variable is color
typeof(diamonds$color) 
# 6. Using summary function tell how many diamonds has ideal cut
summary(diamonds$cut)
# 7. Extract data only for diamonds that cost more than $2401 and store as Expensive
Expensive <- diamonds[diamonds$price > 2401,]
# 8. Using describe function inside psych tell what is mean, standard deviation and range of price for the diamonds.
describe(Expensive$price)
# 9. Find a range of carat for expensive diamonds using basic r
range(Expensive$carat)
# 10. Create a histogram using basic r for the carats using binwith 0.5
hist(Expensive$carat, binw = 0.5)
# 11. Create a ggplot using ggplot2 library for the carats store them as carat
carat <- ggplot(data = Expensive, aes(x=carat))
carat
# 12. Create histogram and add the binwidth the same as we used in basic r
carat + geom_histogram(binwidth = 0.5)
# 13. colour the histogram mapping cut
carat + geom_histogram(binwidth = 0.5, aes(fill = cut))
# 14. set up border for different types in the histogram
carat + geom_histogram(binwidth = 0.5, aes(fill = cut), colour = 'Black')
# 15. Add the main title to the graph
carat + geom_histogram(binwidth = 0.5, aes(fill = cut), colour = 'Black') +
  ggtitle('Carats of expensive diamonds')
# 16. Create a density plot using basic r for the price
plot(density(Expensive$price))
# 17. Create a ggplot using ggplot2 library for the price and store it as price
Price <- ggplot(data = Expensive, aes(x=price))
# 18. Create density plot
Price + geom_density() # create density plot
# 19. Colour the density plot for different cuts
Price + geom_density(aes(fill = cut)) # create density plot for different cuts
# 20. Use position stack to see better density distribution for different cuts
Price + geom_density(aes(fill = cut), position = 'stack')
```


